---
title: "Reproducable Research - Week 2"
author: "Tanmay"
date: "04/07/2021"
output: html_document
---
Load Libraries
```{r}
library(ggplot2)
library(dplyr)
```



***<font size="6"> Loading and preprocessing the data*** 

```{r}
if(!file.exists("./data")){dir.create("./data")}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileUrl,destfile="./data/activity.zip")
unzip(zipfile="./data/activity.zip",exdir="./data")
actdata <- read.csv("./data/activity.csv")
actdata$date <- as.Date(actdata$date)
```


***<font size="6">What is mean total number of steps taken per day? <font size="3">***

1. Calculate the total number of steps taken per day. (top 10)


```{r,echo=TRUE}
stepsByDay <- actdata %>%
        group_by(date) %>%
        summarize(sumsteps = sum(steps, na.rm = TRUE)) 
head(stepsByDay,10)
```


2. Make a histogram of the total number of steps taken each day


```{r,echo=TRUE}
hist(stepsByDay$sumsteps, main = "Histogram of Daily Steps", col="blue", xlab="Steps", ylim = c(0,40))
```


3. Calculate and report the mean and median of the total number of steps taken per day


```{r,echo=TRUE}
meanWithNa <- round(mean(stepsByDay$sumsteps),digits = 2)
medianWithNa <- round(median(stepsByDay$sumsteps),digits = 2)
print(paste("The mean is: ", meanWithNa))
print(paste("The median is: ", medianWithNa))
```


***<font size="6">What is the average daily activity pattern?***


```{r,echo=TRUE}
stepsbyInterval <- actdata %>%
        group_by(interval) %>%
        summarize(meansteps = mean(steps, na.rm = TRUE)) 
#Display first 10 rows of data
head(stepsbyInterval,10)
```



<font size="3"> 1.Make a time series plot (i.e. \color{red}{\verb|type = "l"|}type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r,echo=TRUE}
plot(stepsbyInterval$meansteps ~ stepsbyInterval$interval,
     col="red", type="l", xlab = "5 Minute Intervals", ylab = "Average Number of Steps",
     main = "Steps By 5 Minute Intervals")
```


2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r,echo=TRUE}
print(paste("Interval containing the most steps on average: ",stepsbyInterval$interval[which.max(stepsbyInterval$meansteps)]))
```


***<font size="6">Imputing missing values***


<font size="3">1. Calculated missing values

```{r,echo=TRUE}
print(paste("The total number of rows with NA is: ",sum(is.na(actdata$steps))))
```


2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

Missing Values Strategy: The missing NA values will be equal to the average values calculated in the previous step.

```{r,echo=TRUE}
actNewNaValues <- actdata  
for (i in 1:nrow(actdata)){
        if(is.na(actdata$steps[i])){
                actNewNaValues$steps[i]<- stepsbyInterval$meansteps[actNewNaValues$interval[i] == stepsbyInterval$interval]
        }
}
```


3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r,echo=TRUE}
head(actNewNaValues,10)
```

4.Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

*Table showing new values*

```{r,echo=TRUE}
stepsByDay <- actNewNaValues %>%
        group_by(date) %>%
        summarize(sumsteps = sum(steps, na.rm = TRUE)) 
head(stepsByDay,10)
```

*Histogram showing new values*

```{r,echo=TRUE}
hist(stepsByDay$sumsteps, main = "Histogram of Daily Steps", 
     col="green", xlab="Steps")
```

*Calculate the new mean and median*

```{r,echo=TRUE}
meanWithNewNaValues <- round(mean(stepsByDay$sumsteps), digits = 2)
medianWithNewNaValues <- round(median(stepsByDay$sumsteps), digits = 2)
print(paste("The mean is: ", mean(meanWithNewNaValues)))
print(paste("The median is: ", median(medianWithNewNaValues)))
```

*compare before and after*

```{r,echo=TRUE}
Compare <- data.frame(mean = c(meanWithNa,meanWithNewNaValues),median = c(medianWithNa, medianWithNewNaValues))
rownames(Compare) <- c("With NA Values", "With New NA Values")
print(Compare)
```


***<font size="6">Are there differences in activity patterns between weekdays and weekends?***

<font size="3">1. Create a new factor variable in the dataset with two levels - “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r,echo=TRUE}
activityDoW <- actNewNaValues
activityDoW$date <- as.Date(activityDoW$date)
activityDoW$day <- ifelse(weekdays(activityDoW$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")
activityDoW$day <- as.factor(activityDoW$day)
```

2. Make a panel plot containing a time series plot (i.e. type=“l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r,echo=TRUE}
activityWeekday <- filter(activityDoW, activityDoW$day == "weekday")
activityWeekend <- filter(activityDoW, activityDoW$day == "weekend")
activityWeekday <- activityWeekday %>%
        group_by(interval) %>%
        summarize(steps = mean(steps)) 
activityWeekday$day <- "weekday"
activityWeekend <- activityWeekend %>%
        group_by(interval) %>%
        summarize(steps = mean(steps)) 
activityWeekend$day <- "weekend"
wkdayWkend <- rbind(activityWeekday, activityWeekend)
wkdayWkend$day <- as.factor(wkdayWkend$day)
g <- ggplot (wkdayWkend, aes (interval, steps))
g + geom_line() + facet_grid (day~.) + 
        theme(axis.text = element_text(size = 12),axis.title = element_text(size = 14)) + 
        labs(y = "Number of Steps") + labs(x = "Interval") + 
        ggtitle("Average Number of Steps - Weekday vs. Weekend") + 
        theme(plot.title = element_text(hjust = 0.5))
```